Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Tap

We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models", which accumulate a (potentially infinite) nondeterministic trace of monoidal "events". This will later come in handy for reasoning about output behaviour and weak memory.

@@

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and 
  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
  \TODO{this}

  \TODO{deterministic trace monad; inherited Elgot structure: use as example of "Elgot submonad"? Go cite \cite{carette-central-2022}}
\end{definition}

@@

\TODO{merge det/nondet}

\TODO{induces a substructural effectful category...}

\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}

\TODO{relevant = idempotent}

\TODO{affine = no effect on trace}

\TODO{induces a substructural effectful category...}

\TODO{central = commutative, "commutes with infinity" (i.e. is absorbed by infinity)}

\TODO{relevant = deterministic + idempotent; inclusion of trace monad}

\TODO{affine = no effect on trace}

>>>>

\TODO{fix this text}

We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially infinite) nondeterministic trace of monoidal "events". These will later come in handy for reasoning about output behaviour and weak memory.

@@

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and 
  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
  \TODO{this}
\end{definition}

@@

\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
\end{definition}
Some important examples of submonads include:
\begin{itemize}
  \item \todo{unit submonad of every monad}
  \item \todo{union, intersection of (Elgot) submonads; lattice of submonads}
  \item \todo{commuting absorbtive submonad of trace monad: example of...}
  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
  \item \todo{deterministic, idempotent trace monad: relevant submonad}
  \item \todo{noneffectful trace monad: affine submonad}
  \item \todo{cap and centralize, cap and purify}
\end{itemize}
\TODO{in particular, these induce a substructural effectful category...}
